Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-14 paper/tidy-missing-data-paper.Rmd
minor text amendments
minor changes to figures + some titles + reference later sections in the paper.

library(tidyverse)
library(gridExtra)
# options(kableExtra.latex.load_packages = FALSE)
# library(kableExtra)
library(ggthemes)
library(png)
library(scales)
library(simputation)
# if paper doesn't compile, use development version of naniar
# install.packages("remotes")
# remotes::install_github("njtierney/naniar")
library(naniar) 
library(knitr)
library(lubridate)
library(magick)
options(tinytex.verbose = TRUE)
```

@@

The imputation literature (e.g. @Little1988, @Rubin1976, @Simon1986,
@Schafer2002, @VanBuuren2012) focusses on ensuring valid statistical inference
is made from incomplete data. This is approached chiefly through probabilistic
modelling, assuming the mechanism of missing data is known to the analyst. It
does not address how to explore, understand, and handle missing data structures
and mechanisms.

@@

While there have been many software tools for exploring missing data [@VIM, @mi, @mice, @Cheng2015, @MANET, @ggobi, @norm], these tools do not integrate easily into analysis pipelines. Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline. The analyst must
explore missing data using visualizations, summaries, and exploratory analyses,
but there is a lack of modern tools to efficiently explore missing data.

The graphics literature [@Swayne1998; @Unwin1996; @VIM; @Cheng2015] provides several solutions for exploring missings visually. These approaches incorporate missings into the plot in some way. For example, imputing values 10% below the range to display all values in a scatter plot. These methods require the data to be linked to the missing or imputed values, but this link is typically hidden from the user. The approaches from the graphics literature have not been translated and integrated into tidy data. This limits the analyst's capacity to efficiently explore missing values. Extending tidy data and tidy tools to account for missing values would create more efficient and robust workflows and analysis.

@@

# Extensions to tidyverse {#extensions}

@@

library(visdat)

@@

```{r parallel-cord-plot, fig.width = 8, fig.height = 4, out.width = "100%", echo = FALSE, fig.cap = "Parallel coordinate plot shows missing values imputed 10\\% below range for the oceanbuoys dataset. Values are colored by missingness of humidity. Humidity is missing for low air and sea temperatures, and is missing for one year and one location. "}
library(naniar)
library(tidyverse)

@@

Figure \@ref(fig:housing-miss-case-var)A shows 9 variables with missing values. The most missings are in building area, followed by year built, and land size, with similar amounts of missingness in Car, bathroom, bedroom2, longitude, and latitude. Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing. The variables building area and year built have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious.

@@

Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).

@@

```{r rpart-fit-cluster}
library(rpart)

@@

aug_fits %>%
  ggplot(aes(x = .fitted,
             y = .resid)) + 
  # geom_point(alpha = 0.1) + 
  geom_hex() +
  labs(fill = "Count") +
  scale_fill_viridis_c() +
  facet_wrap(~model)

>>>>

library(tidyverse) 
library(gridExtra)
# options(kableExtra.latex.load_packages = FALSE)
# library(kableExtra)
library(ggthemes)
library(png)
library(scales)
library(simputation)
# if paper doesn't compile, use development version of naniar
# install.packages("remotes")
# remotes::install_github("njtierney/naniar")
library(naniar) 
library(knitr)
library(lubridate)
library(magick)
library(rpart)
library(visdat)
library(patchwork)
options(tinytex.verbose = TRUE)
```

@@

The imputation literature (See \@ref(imputation)) focusses on ensuring valid statistical inference
is made from incomplete data. This is approached chiefly through probabilistic
modelling, assuming the mechanism of missing data is known to the analyst. It
does not address how to explore, understand, and handle missing data structures
and mechanisms.

@@

While there have been many software tools for exploring missing data (See \@ref(exploration)), these tools do not integrate easily into analysis pipelines. Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline. The analyst must
explore missing data using visualizations, summaries, and exploratory analyses,
but there is a lack of modern tools to efficiently explore missing data.

The graphics literature (See \@ref(exploration)) provides several solutions for exploring missings visually. These approaches incorporate missings into the plot in some way. For example, imputing values 10% below the range to display all values in a scatter plot. These methods require the data to be linked to the missing or imputed values, but this link is typically hidden from the user. The approaches from the graphics literature have not been translated and integrated into tidy data. This limits the analyst's capacity to efficiently explore missing values. Extending tidy data and tidy tools to account for missing values would create more efficient and robust workflows and analysis.

@@

# Tidy framework for Missings {#extensions}

@@

@@

```{r parallel-cord-plot, fig.width = 8, fig.height = 4, out.width = "100%", echo = FALSE, fig.cap = "Parallel coordinate plot shows missing values imputed 10\\% below range for the oceanbuoys dataset. Values are colored by missingness of humidity. Humidity is missing for low air and sea temperatures, and is missing for one year and one location. "}

@@

Figure \@ref(fig:housing-miss-case-var)A shows 9 variables with missing values. The most missings are in building area, followed by year built, and land size, with similar amounts of missingness in Car, bathroom, bedroom2, longitude, and latitude. Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing. The variables building area and year built have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious. Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).

@@

@@

```{r rpart-fit-cluster}

@@

gg_hex <- function(data,
                   subset){

  data_subset <- data %>% 
    filter(model == subset)

    ggplot(data_subset, aes(x = .fitted,
               y = .resid)) + 
    geom_hex() + 
    scale_fill_viridis_c() +
    theme(legend.position = "bottom",
          aspect.ratio = 1) +
      labs(tag = subset)
}

gg_hex(aug_fits, "cc") +
gg_hex(aug_fits, "knn") +
gg_hex(aug_fits, "lm")
